### 2019-10-01 ch1_introduction/content.tex
update

Developing new computational tools to perform these analyses has proven challenging for two main reasons \cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}. Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity. In addition, a low abundance of single-cell omics data that can be used as a ground-truth obstruct quantitative evaluation of such methods.

@@

\subsection{Gene regulatory network inference}
Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes. This topic already received much attention with the advent of bulk omics (before single-cell omics). These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods\cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.

@@

\begin{itemize}
  \item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
  \item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
  \item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
  \item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
  \item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
  \item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}). We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
  \item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
  \item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
  \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. bred: Inferring single cell regulatory networks. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
 \end{itemize}

>>>>

Developing new computational tools to perform these analyses has proven challenging for two main reasons\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}. Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity. In addition, a low abundance of single-cell omics data that can be used as a ground-truth obstruct quantitative evaluation of such methods.

@@

\subsection{Network inference}
Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes. This topic already received much attention with the advent of bulk omics (before single-cell omics). These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods\cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.

@@

\begin{itemize}
  \item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
  \item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
  \item We developed dyno, a toolkit to easily infer, visualise and interpret single-cell trajectories using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
  \item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
  \item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
  \item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
  \item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}). We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
  \item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
  \item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
  \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. \fbox{\color{red}order authors?}
  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
 \end{itemize}
